Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-09-08 paper/paper.tex
Danilos revision 1 for NHB

In the present manuscript, we propose a \textit{process model}
that attempts to explain \textit{how} the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
We first review the rich literature on the DMN and recognize that the DMN contains the sufficient neuro-biological substrates to implement a full-blown reinforcement-learning agent, at least in principle. DMN function is then recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{Averbeck2017,amygdala_bzdok}.

@@

Consequently, the PMC may contribute to overall DMN function by
``altering behavior in response to unexpected change''
\citep{pearson2011posterior}.
This DMN node probably
monitors sudden and gradual environmental changes in subjective outcomes
of action choices for integration
with memory and perspective frames
to improve short- and longer-term behavioral agendas.
Estimated value, that differs across individuals, probably enriches
statistical assessment of the environment
to map and predict delayed reward opportunities in the future.
In doing so, the PMC may continuously adapt the organism to changes
in both the external environment and its internal representation
to enable strategic behavior.

@@

RL is a problem-solving technique in which,
through repeated interactions with an \textit{environment},
an agent learns to reach goals and optimize reward signals occasionally emitted by the environment, in an iterative trial-and-error fashion (Fig. \ref{fig:rl}).
At a given moment, each taken \textit{action} $a$ triggers a change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a,s')$ collected by the agent.
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction \textemdash or aversion \textemdash accompanying the execution of
a particular action.  Thus the agent learns behaviors.
% Action may be delayed to achieve substantial increases in expected reward
% that is expected to grow with time.

\begin{wrapfigure}{r}{.35\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception~\citep{starkweather2017dopamine,o2006making,o2006making}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\subsection{Interlude: Rethinking the Action-State-Reward model of RL}
Barto proposes to decouple extrinsic rewards (i.e generated by the external environment) from intrinsic reward (generated from withing the "organism")
biologically, extrinsic rewards would correspond to so-called "sensations"
while intrinsic rewards would correspond to "reward signals".

Thus there are two environments: the external environment, and the internal environment built-into the "organism". The organism takes highly level DECISIONS (the DMN can confortably sit here!) which act on the internal environment (this environment probably corresponds to low-level subsystems like motor, etc.) which then transform theses  into ACTIONS executed in the external environment. The external environment generates sensations which get mapped into reward signals (via the internal environment), etc.

The interesting thing is that with this rethinking, the organism can have a perception-action loop even in the absence of external stimulus.
This is called curiosity...

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games like the breakthrough results upon recently on the game of Go \citep{silver2016mastering}, considered to be a golden benchmark problem in artificial intelligence.
In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in active learning.

@@

\begin{mdframed}
\paragraph*{Why MDPS ?} At this point, it is righteous to question the applicability of MPDs as a model for something as complex as human behavior ?
Indeed a critique could argue that Markov chains are state-machines without memory. These may describe insects, or other behaviorally primitive beings like ATM machines, insects, etc., but won't be sufficient to describe human beings, as for example, human behavior is heavily influenced not only by the current state but by past states as well (memory).

\paragraph*{Answer:} We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.

Rational probabilistic planning can be reformulated
as a standard memory Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.
This adds the capacity for memory, since then, the next state depends not only on the current state but also on a finite number of past states. This is precisely what partially observable MDPs do ~\citep{starkweather2017dopamine,o2006making,o2006making}.

Also, it should be noted that MDPs have been successfully used as neural model for animal behavior. One can mention ~\citep{Rao2010}, just to name a few.
\end{mdframed}

@@

Optimal behavior of the agent corresponds to
a strategy $\pi^*$ for choosing actions such that, for every state, the chosen action guarantees the best possible reward on average. Formally,
% \textbf{XXX please add 1-2 sentences why this is interesting in the current
% context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) := \argmax_{a \in \mathcal A}{Q}^*(s, a),\text{ where }Q^*(s, a) := \max_{\pi}Q^\pi(s,a).
  \label{eq:qlearning}
\end{equation}
The learning goal is to approach the
policy $\pi^*$ as close as possible, that is to \text{solve} the MDP.
Note that \eqref{eq:qlearning} presents merely a definition and
does not lend itself as a candidate schema
for solving MDPs with even moderately-sized action and state spaces
(i.e., intractability).
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, with the form
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the so-called Bellman transform $\text{Bel}(Q)$
of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is another $Q$-value function defined by
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a temporal consistency equation which provides
a dynamic decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
The optimal Q-value operator $Q^*$ is a fixed point for this equation.
As a consequence of this decomposition, the complicated problem \eqref{eq:qlearning}
is broken down into simpler sub-problems at different time points.
Indeed, exploitation of hierarchical structure in action considerations
has previously been related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a preferred state, thus improving on the 
current action policy of the agent.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the random sampling
is performed only over quantities which
depend on the environment.
This aspect of the learning process
can unroll off-policy by observing state transitions
triggered by another (possibly stochastic) behavioral policy.

\pagebreak[4]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Neural correlates of the Bellman equation in the DMN:}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN node were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics \citep{lavilleon2015}.
This would be a simple control algorithm by evaluating all legal actions and selecting the action with
highest expected cumulative rewards.As a result, quick-to-learn instance-based control policies serve as a rough approximation while a slower more generalisable decision-making system system is trained up~\citep{pritzel2017neural,lengyel2008hippocampal}.
In MDPs, MCMC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to theories of the free-energy principle (FEP) and
active inference \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical reasoning engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this interpretation can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also proposes a companion model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensory sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free energy} $F_G^R(\x) := \langle \log p_G(., \x) \rangle_{p_R(. |\x)} - \langle \log p_R(. |\x))\rangle_{p_R(. |\x)}$ given by

  \begin{equation}
    \begin{split}
      \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
      &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(P_R(.|\x) || P_G(.|\x))}_{\text{complexity}} \\
      &\le F^R_G(\x),
      \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z.
    \end{split}
  \end{equation}

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm is that it involves a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
The brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
The recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free energy, VAEs can be efficiently trained via back-propagation of prediction errors. Most importantly, the reparametrization trick allows for a drastic reduction in the variance of the in estimates of the model's gradients, leading to effective learning, something which was not possible in wake-sleep\citep{dayan1995helmholtz}. The details of this trick can be found in \citep{kingma2013auto}.

@@

\appendix

>>>>

The present paper proposes a \textit{process model}
that tries to explain \textit{how}
the DMN may implement
continuous evaluation and prediction of the environment to guide behavior.
%
DMN function is recast in mathematical terms of
control theory and reinforcement learning based on Markov decision processes.
%
We argue that our formal account of DMN function naturally accommodates as special cases
the previously proposed cognitive accounts on
(1) predictive coding,
(2) semantic associations, and
(3) a ``sentinel'' role.
%
Moreover, this process model for the neural optimization of complex behavior in the DMN
offers parsimonious explanations for
recent experimental findings in animals and humans.
\end{abstract}

@@

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.
We also contribute to the discussion of DMN function
by providing some of the first empirical evidence that
morphological variability in DMN nodes
is linked to the reward circuitry (Fig. 2).
Finally, we detail how our process model relates
to previous accounts of DMN function and
we derive explicit hypotheses to be tested in future neuroscience
experiments (Box 1).

\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
\textbf{BOX 1: Hypotheses for testing the MDP account of DMN function}

\begin{enumerate}
\item Experiment (Humans): We hypothesize a functional relationship between the DMN closely associated with the occurrence of stimulus-independent thoughts and the reward circuitry. During an iterative neuroeconomic two-player game, fMRI signals in the DMN could be used to predict reward-related signals in the nucleus accumbens across trials in a continuous learning paradigms. We expect that the more DMN activity is measured to be increased, supposedly the higher the tendency for stimulus-independent thoughts, and the more the fMRI signals in the reward circuits should be independent of the reward context
in the sensory environment.

\item Experiment (Humans): We hypothesize a functional dissociation between computations for behavioral policy versus adapting stimulus-value associations as these are probably implemented in different subsystems of the DMN. We first expect that fMRI signals in the right temporo-parietal junction can predict behavioral changes caused by adaptation in the action choice tendencies (policy matrix) related to non-value-related prediction error. Second, fMRI signals in the ventromedial prefrontal cortex should predict behavioral changes caused by adaptation in value estimation (value matrix) due to reward-related stimulus-value association. We finally predict that fMRI signals in the posteromedial cortex, as a potential global information integrator, is able to predict shifts in overt behavior based on previous adaptations in policy or value estimation.

\item Experiment (Animals): We hypothesize that experience replay for browsing problem solutions subserved by the DMN is necessary for choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before make analogous choices in the future. Necessity of the DMN, in addition to the hippocampus, for “mind-searching” actions during choice behavior can be demonstrated by causal dysruption of DMN nodes, such as circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

\end{enumerate}
\end{mdframed}

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Neurological patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological relevance assessment
\citep{amygdala_bzdok}.

@@

Consequently, the PMC may contribute to overall DMN function
by monitoring the subjective outcomes
of possible actions and integrating that information
with memory and perspective frames
into short- and longer-term behavioral agendas.
Estimated value, that differs across individuals, probably enriches
statistical assessment of the environment
to map and predict delayed reward opportunities in the future.
In doing so, the PMC may continuously adapt the organism to changes
in both the external environment and its internal representation
to enable strategic behavior.

@@

RL is a problem-solving technique in which,
through repeated interactions with an \textit{environment},
an agent learns to reach goals and optimize reward signals
in an iterative trial-and-error fashion (Fig. \ref{fig:rl}).
At a given moment, each taken \textit{action} $a$ triggers a change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a,s')$ collected by the agent.
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction \textemdash or aversion \textemdash accompanying the execution of
a particular action.
% Action may be delayed to achieve substantial increases in expected reward
% that is expected to grow with time.

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensory perception.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

\begin{wrapfigure}{r}{.3\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

@@

\subsection{Markov decision processes}
Model-free RL has had great success in many real-world problems,
including robotics \citep{ng2004,abbeel2004},
super-human performance in complex video games \citep{mnih2015},
and strategic board games like the breakthrough results upon recently on the game of Go \citep{silver2016mastering}, considered to be a golden benchmark problem in artificial intelligence.
We emphasize that the brain in general, and the DMN in particular,
is a physical system governed by the laws of
physics and can be formally described
by Markov processes at a sufficiently coarse scale.
It has indeed been previously proposed \citep{tegmark2016improved} that
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short.
% , defining $s(t)$, the state at time $t$, as the position in state space.
The process has $memory$ if the next state depends not only on the current state
but also on a finite number of past states.
Rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include experience episodes of the past.

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state),
by maintaining probability distributions over actions according to their expected utility.
% This is performed alongside each transition in the state space of the environment
%  and maintains a probability distribution over the set of possible
% states.
% % , based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment faced by an agent engaged in active learning.

@@

@@

Optimal behavior of the agent corresponds to
a strategy $\pi^*$ for choosing actions such that, for every state, the chosen action guarantees the best possible reward on average. Formally,
% \textbf{XXX please add 1-2 sentences why this is interesting in the current
% context. This statement appears a little alone as it stands.}
\begin{equation}
  \pi^*(s) := \argmax_{a \in \mathcal A}{Q}^*(s, a),\text{ where }Q^*(s, a) := \max_{\pi}Q^\pi(s,a).
  \label{eq:qlearning}
\end{equation}
The learning goal is to approach the
policy $\pi^*$ as close as possible, that is to \text{solve} the MDP.
Note that \eqref{eq:qlearning} presents merely a definition and
does not lend itself as a candidate schema
for solving MDPs with even moderately-sized action and state spaces
(i.e., intractability).
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
Fortunately, the \textit{Bellman equation} \citep{sutton1998reinforcement} provides a fixed-point relation which defines $Q^*$ implicitly via a sampling procedure, without querying the entire space of policies, with the form
\begin{equation}
  Q^* = \text{Bel}(Q^*),
  \label{eq:bellman}
\end{equation}
where the so-called Bellman transform $\text{Bel}(Q)$
of an arbitrary $Q$-value function $Q: \mathcal S \times \mathcal A \rightarrow \mathbb R$  is another $Q$-value function defined by
\begin{equation}
  \begin{split}
   \text{Bel}(Q)(s,a) &:=
   \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q(s', a')]\\
   &= r(s,a) + \gamma\mathbb E_{s' \sim p(s'|s,a)} [\max_{a' \in \mathcal A}Q(s', a')]\\
   &= \text{instantaneous reward} + \text{expected reward for acting greedily thereafter}
%    \; \forall (s,a) \in \mathcal S \times \mathcal A
    % \\&=
    % \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^*(s', \pi^*(s'))].
    \end{split}
  \end{equation}
The Bellman equation \eqref{eq:bellman} is a temporal consistency equation which provides
a dynamic decomposition of optimal behavior by dividing the $Q$-value function into the immediate
reward and the discounted rewards of the upcoming states.
The optimal Q-value operator $Q^*$ is a fixed point for this equation.
As a consequence of this decomposition, the complicated dynamic programming
problem \eqref{eq:qlearning}
is broken down into simpler sub-problems at different time points.
Indeed,
exploitation of hierarchical structure in action considerations
has previously been related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value
to guide action towards a preferred state, thus improving on the 
current action policy of the agent.
% It is a fixed-point equation for the deterministic policy $\pi$.
Note that in \eqref{eq:bellman} the random sampling
is performed only over quantities which
depend on the environment.
This aspect of the learning process
can unroll off-policy by observing state transitions
triggered by another (possibly stochastic) behavioral policy.

\pagebreak[4]
\begin{mdframed}
\hspace{2mm}
\vspace{2mm}
% neuroscience of Bellman equation
\textbf{Box 2: Neural correlates of the Bellman equation in the DMN}
Relating decomposition of consecutive action choices by the Bellman equation
to neuroscience,
specific neural activity in the dorsal prefrontal cortex (BA9)
was linked to processing ``goal-tree sequences''
in human neuroimaging experiments
\citep{koechlin1999role, koechlin2000dissociating}.
Sub-goal exploration may require multi-task switching
between cognitive processes
as later parts of a solution
frequently depend on respective earlier steps in a given solution path, which
necessitates storage of expected intermediate outcomes.
As such,
``cognitive branching'' operations for nested processing of behavioral strategies
are likely to
entail secondary reallocation of attention and working-memory resources.
Further neuroimaging experiments \citep{braver2002role} corroborated
the prefrontal DMN to subserve
``processes related to the management and monitoring of sub-goals while maintaining information in working memory''.
%
Moreover,
neurological patients with lesions in this DMN node were reported
to be impaired in aspects of realizing ``multiple sub-goal scheduling''
\citep{burgess2000cognitive}.
Hence,
the various advanced human mental abilities subserved by the DMN, such as
planning and abstract reasoning, can be viewed to involve some form of
action-decision branching to enable higher-order executive control.
\end{mdframed}

@@

In the human brain,
the HC could contribute to synthesizing imagined sequences of world states,
actions and rewards \citep{aronov2017, chao2017interaction, boyer2008evolutionary}.
These simulations of experience batches
would be used to update the value function, without ever looking inside the black box describing the model's dynamics \citep{lavilleon2015}.
This would be a simple control algorithm by evaluating all legal actions and selecting the action with
highest expected cumulative rewards.
In MDPs, MCMC simulation provides an effective mechanism both for tree search and for belief-based state updates, breaking the curse of dimensionality and allowing much greater scalability than has previously been possible \citep{silver2016mastering}.
% Much recent work points to MCMC or stochastic sampling-based approximations as a unifying framework for understanding how Bayesian inference may work practically across all these levels, in minds, brains, and machines. \textbf{XXX elvis: refs needed in last sentence. Plus, should be really bring up Bayesian?}

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to theories of the free-energy principle (FEP) and
active inference \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical reasoning engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this interpretation can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ of objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also proposes a companion model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensory sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free energy} $F_G^R(\x)$ given by

  \begin{equation}
    \begin{split}
      \text{generative surprise } &:= -\log(p_G(\x)) = F_G(\x) \\
      &=\underbrace{F^R_G(\x)}_{\text{accuracy}} - \underbrace{D_{KL}(P_R(.|\x) || P_G(.|\x))}_{\text{complexity}} \\
      &\le F^R_G(\x),
      \text{ with equality if }p_R(\z|\x) = p_G(\z|\x)\text{ for all } \z
    \end{split}
  \end{equation}

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which outlined in the following.
The main algorithm for minimizing free energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As these authors noted,
a crucial drawback of the wake-sleep algorithm is that it involves a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
The brain may therefore be unlikely to implement a variant of the wake-sleep algorithm.
The recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free energy, VAEs can be efficiently trained via back-propagation of prediction errors.

@@

% \appendix
% 
% \begin{remark}
%   Given a system $(\mathcal S, \mathcal A, r, p, \mu)$ satisfying all the axioms for an MDP
